### 2018-05-16 submission/manuscript.Rmd
remove extraction kit from abstract
update intro 2 paragraph --remove emphasis on preservation and extraction

**Background.** Bias is introduced at different stages of the 16S rRNA gene sequencing workflow. Commonly studied sources of bias include preservation and DNA extraction methods. Although the number of cycles and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow. Here, we examine how both cycle number and polymerase can change the bacterial community and introduce bias to the final obtained results.

@@

Accounting for bias is critical in order to reach an understanding of bacterial community changes using 16S rRNA gene sequencing results. Differentiating between bias, reproducibility, and standardization is important since often times these three can be confused and used interchangeably with each other. Bias can change the observed results in a way that is reproducible and standardized. For example, if one group uses one brand of DNA extraction kit for their 16S rRNA gene sequencing, their results may be biased versus another group not using the same brand kit but within their group they can still have reproducible results [@contamination_Salter2014]. Our 16S rRNA gene sequencing methods are biased even when these workflows are standardized to increase reproducibility. In order to interpret specific studies within the broader context of the overall field, assessing bias at different parts of the 16S rRNA gene sequencing workflow is critical.

Many parts of the 16S rRNA gene sequencing workflow contribute bias to the results and are studied extensively. A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. Generally, not using a preservation media and leaving samples at room temperature supports overgrowth of low abundance members of the fecal bacterial community [@Amir2017]. Similarly, this overgrowth can still occur if the preservation media does not adequately inhibit growth [@preservation_Luo_2016; @preservation_Song_2016]. Reports have also shown that changes in specific community members might occur due to differing susceptibility to freeze thaw cycles amongst microbes [@Gorzelak2015]. Additionally, reagent contamination can add community members and the contribution of these contaminant members grows larger with lower biomass samples [@contamination_Salter2014]. Overall, biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [@preservation_Song_2016; @storage_Bassis_2017]. However, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method.

@@

The intrinsic properties to primers and DNA polymerases chosen can also introduce bias. Primers have variable region dependent binding affinities for different bacteria and depending on the primer pair do not detect specific bacteria (e.g. V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)]. Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capacity [@polymerase_Ishino_2014]. Interestingly, the influence that these different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other previously mentioned sources of PCR-based bias.

A recent study found clear differences between normal and high fidelity DNA polymerase and that optimization of the PCR protocol could reduce error and chimera generation [@taq_Gohl_2016]. This study also found that regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) or taxa generated were not easily reduced using the authors chosen bioinformatic pipeline [@taq_Gohl_2016]. It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on the type of high fidelity DNA polymerase. There is some reason to think that this may be the case since many of these HiFi DNA polymerases come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014].

Although bias introduced due to differences between DNA polymerase and high fidelity DNA polymerase has been investigated for 16S rRNA gene sequencing, the bias caused due to differences between specific high fidelity DNA polymerases has not been. This study will specifically address how high fidelity DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing. We will accomplish this by examining if any of five different types of high fidelity DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

>>>>

**Background.** Bias is introduced at different stages of the 16S rRNA gene sequencing workflow. Although the number of cycles and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow. Here, we examine how both cycle number and polymerase can change the bacterial community and introduce bias to the final obtained results.

@@

Accounting for bias is critical in order to reach an understanding of bacterial community changes using 16S rRNA gene sequencing results. Bias can change the observed results in a way that is reproducible and standardized. Within this context, our 16S rRNA gene sequencing methods are biased even when these workflows are standardized to increase reproducibility. In order to interpret specific studies within the broader context of the overall field, assessing bias at different parts of the 16S rRNA gene sequencing workflow is critical.

Many parts of the 16S rRNA gene sequencing workflow contribute bias to the results and are studied extensively. A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. Overall, biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [@preservation_Song_2016; @storage_Bassis_2017]. Although the observed biological signal is larger, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method.

@@

The intrinsic properties to primers and DNA polymerases chosen can also introduce bias. Primers have variable region dependent binding affinities for different bacteria and depending on the primer pair do not detect specific bacteria (e.g. V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)]. Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capacity [@polymerase_Ishino_2014]. Interestingly, the influence that these different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other sources of PCR-based bias.

A recent study found clear differences between normal and high fidelity DNA polymerase and that optimization of the PCR protocol could reduce error and chimera generation [@taq_Gohl_2016]. This study also found that regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) or taxa generated were not easily reduced using the authors chosen bioinformatic pipeline [@taq_Gohl_2016]. It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on the type of high fidelity DNA polymerase. There is some reason to think that this may be the case since many of these high fidelity DNA polymerases come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014].

Although bias introduced due to differences between DNA polymerase and high fidelity DNA polymerase has been investigated for 16S rRNA gene sequencing, the bias caused due to differences between specific high fidelity DNA polymerases has not been. This study will investigate how high fidelity DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing. We will accomplish this by examining if any of five different types of high fidelity DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

### 2018-05-16 submission/manuscript.tex
remove extraction kit from abstract
update intro 2 paragraph --remove emphasis on preservation and extraction

\textbf{Background.} Bias is introduced at different stages of the 16S
rRNA gene sequencing workflow. Commonly studied sources of bias include
preservation and DNA extraction methods. Although the number of cycles
and high fidelity (HiFi) DNA polymerase are studied less often, they are
still important sources of bias in this workflow. Here, we examine how
both cycle number and polymerase can change the bacterial community and
introduce bias to the final obtained results.

@@

Accounting for bias is critical in order to reach an understanding of
bacterial community changes using 16S rRNA gene sequencing results.
Differentiating between bias, reproducibility, and standardization is
important since often times these three can be confused and used
interchangeably with each other. Bias can change the observed results in
a way that is reproducible and standardized. For example, if one group
uses one brand of DNA extraction kit for their 16S rRNA gene sequencing,
their results may be biased versus another group not using the same
brand kit but within their group they can still have reproducible
results (Salter et al., 2014). Our 16S rRNA gene sequencing methods are
biased even when these workflows are standardized to increase
reproducibility. In order to interpret specific studies within the
broader context of the overall field, assessing bias at different parts
of the 16S rRNA gene sequencing workflow is critical.

Many parts of the 16S rRNA gene sequencing workflow contribute bias to
the results and are studied extensively. A typical 16S rRNA gene
sequencing workflow can be divided into preservation, extraction, PCR,
and sequencing steps. Generally, not using a preservation media and
leaving samples at room temperature supports overgrowth of low abundance
members of the fecal bacterial community (Amir et al., 2017). Similarly,
this overgrowth can still occur if the preservation media does not
adequately inhibit growth (Song et al., 2016; Luo et al., 2016). Reports
have also shown that changes in specific community members might occur
due to differing susceptibility to freeze thaw cycles amongst microbes
(Gorzelak et al., 2015). Additionally, reagent contamination can add
community members and the contribution of these contaminant members
grows larger with lower biomass samples (Salter et al., 2014). Overall,
biases due to either preservation or extraction tend to be smaller than
the overall biological signal being measured (Song et al., 2016; Bassis
et al., 2017). However, the contribution of PCR bias to this overall
workflow is not well characterized since these studies use the same PCR
approach while varying preservation or extraction method.

@@

The intrinsic properties to primers and DNA polymerases chosen can also
introduce bias. Primers have variable region dependent binding
affinities for different bacteria and depending on the primer pair do
not detect specific bacteria (e.g.~V1-V3 does not detect
\emph{Haemophilus influenzae} and V3-V5 does not detect
\emph{Propionibacterium acnes}) (Sze et al., 2015 (Table S4); Meisel et
al., 2016). Additionally, there are multiple families of DNA polymerases
that have their own error rate and proof reading capacity (Ishino \&
Ishino, 2014). Interestingly, the influence that these different DNA
polymerases can have on the observed 16S rRNA gene sequencing results
have not been well studied like some of the other previously mentioned
sources of PCR-based bias.

A recent study found clear differences between normal and high fidelity
DNA polymerase and that optimization of the PCR protocol could reduce
error and chimera generation (Gohl et al., 2016). This study also found
that regardless of DNA polymerase, the number of Operational Taxonomic
Units (OTUs) or taxa generated were not easily reduced using the authors
chosen bioinformatic pipeline (Gohl et al., 2016). It is natural to
extend this line of inquiry and ask if biases in the number of OTUs and
chimeras are also dependent on the type of high fidelity DNA polymerase.
There is some reason to think that this may be the case since many of
these HiFi DNA polymerases come from different families (e.g. \emph{Taq}
belongs to the family A polymerases) and may intrinsically have
different error rates that cannot be completely removed with
modifications (Ishino \& Ishino, 2014).

Although bias introduced due to differences between DNA polymerase and
high fidelity DNA polymerase has been investigated for 16S rRNA gene
sequencing, the bias caused due to differences between specific high
fidelity DNA polymerases has not been. This study will specifically
address how high fidelity DNA polymerases can bias observed bacterial
community results derived from 16S rRNA gene sequencing. We will
accomplish this by examining if any of five different types of high
fidelity DNA polymerases introduce significant biases into 16S rRNA gene
surveys, if this is a cycle dependent phenomenon, and whether they can
be removed using a standard bioinformatic pipeline.

@@

\hypertarget{ref-Gorzelak2015}{}
Gorzelak MA., Gill SK., Tasnim N., Ahmadi-Vand Z., Jay M., Gibson DL.
2015. Methods for improving human gut microbiome data by reducing
variability through sample processing and storage of stool. \emph{PLOS
ONE} 10:e0134802. DOI:
\href{https://doi.org/10.1371/journal.pone.0134802}{10.1371/journal.pone.0134802}.

>>>>

\textbf{Background.} Bias is introduced at different stages of the 16S
rRNA gene sequencing workflow. Although the number of cycles and high
fidelity (HiFi) DNA polymerase are studied less often, they are still
important sources of bias in this workflow. Here, we examine how both
cycle number and polymerase can change the bacterial community and
introduce bias to the final obtained results.

@@

Accounting for bias is critical in order to reach an understanding of
bacterial community changes using 16S rRNA gene sequencing results. Bias
can change the observed results in a way that is reproducible and
standardized. Within this context, our 16S rRNA gene sequencing methods
are biased even when these workflows are standardized to increase
reproducibility. In order to interpret specific studies within the
broader context of the overall field, assessing bias at different parts
of the 16S rRNA gene sequencing workflow is critical.

Many parts of the 16S rRNA gene sequencing workflow contribute bias to
the results and are studied extensively. A typical 16S rRNA gene
sequencing workflow can be divided into preservation, extraction, PCR,
and sequencing steps. Overall, biases due to either preservation or
extraction tend to be smaller than the overall biological signal being
measured (Song et al., 2016; Bassis et al., 2017). Although the observed
biological signal is larger, the contribution of PCR bias to this
overall workflow is not well characterized since these studies use the
same PCR approach while varying preservation or extraction method.

@@

The intrinsic properties to primers and DNA polymerases chosen can also
introduce bias. Primers have variable region dependent binding
affinities for different bacteria and depending on the primer pair do
not detect specific bacteria (e.g.~V1-V3 does not detect
\emph{Haemophilus influenzae} and V3-V5 does not detect
\emph{Propionibacterium acnes}) (Sze et al., 2015 (Table S4); Meisel et
al., 2016). Additionally, there are multiple families of DNA polymerases
that have their own error rate and proof reading capacity (Ishino \&
Ishino, 2014). Interestingly, the influence that these different DNA
polymerases can have on the observed 16S rRNA gene sequencing results
have not been well studied like some of the other sources of PCR-based
bias.

A recent study found clear differences between normal and high fidelity
DNA polymerase and that optimization of the PCR protocol could reduce
error and chimera generation (Gohl et al., 2016). This study also found
that regardless of DNA polymerase, the number of Operational Taxonomic
Units (OTUs) or taxa generated were not easily reduced using the authors
chosen bioinformatic pipeline (Gohl et al., 2016). It is natural to
extend this line of inquiry and ask if biases in the number of OTUs and
chimeras are also dependent on the type of high fidelity DNA polymerase.
There is some reason to think that this may be the case since many of
these high fidelity DNA polymerases come from different families (e.g.
\emph{Taq} belongs to the family A polymerases) and may intrinsically
have different error rates that cannot be completely removed with
modifications (Ishino \& Ishino, 2014).

Although bias introduced due to differences between DNA polymerase and
high fidelity DNA polymerase has been investigated for 16S rRNA gene
sequencing, the bias caused due to differences between specific high
fidelity DNA polymerases has not been. This study will investigate how
high fidelity DNA polymerases can bias observed bacterial community
results derived from 16S rRNA gene sequencing. We will accomplish this
by examining if any of five different types of high fidelity DNA
polymerases introduce significant biases into 16S rRNA gene surveys, if
this is a cycle dependent phenomenon, and whether they can be removed
using a standard bioinformatic pipeline.

@@
